FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories

### 2024-12-04 manuscript.Rmd
update 2024-12-04 17:35:40.415285

abstract: |
  Test test.
keywords          : "meta theory, theory formation, cumulative science, formal models"
wordcount         : "`r tryCatch(wordcountaddin::word_count(here::here('manuscript.Rmd')))`"
bibliography      : "theory-specification.bib"
floatsintext      : no
linenumbers       : yes
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
---

@@

The FAIR Guiding Principles (hereafter: FAIR principles) were established to make research data more Findable, Accessible, Interoperable and Reusable [REF].
Since their inception, scholars have demonstrated their relevance for making other digital research artefacts more open.
This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory as well.
To this end, we introduce "FAIR theory":
a digital representation of theory compliant with the FAIR principles.
FAIR Theory has the potential to substantially advance the efficiency of scholarly communication and accelerate cumulative knowledge acquisition.

@@

Scholars have raised concerns about the state of theory in social science for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
One concern is that social scientific theories often lack the precision and clarity of theories in the physical sciences, otherwise known as *formalization* [@szollosiArrestedTheoryDevelopment2021].
A second concern, which received less attention, is the lack of transparent and democratic scholarly communication about psychological theory.
The present paper focuses on the second concern,
advancing transparent communication about theory by applying open science principles to psychological theory for the first time
and introducing the concept of *FAIR Theory*.

@@

In a progressive research program [@lakatosHistoryScienceIts1971],
the cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
There are indications that contemporary psychology falls short of this idealized model, however.
Firstly, because deductive research is over-represented in the literature: 
According to one estimate, 89.6% of paper published in psychology tests hypotheses [@kuhbergerPublicationBiasPsychology2014].
Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
Only 15% of deductive studies referenced any theory, and theory was often not references in connection to the hypothesis [@mcphetresDecadeTheoryReflected2021].
This invites the question where hypotheses come from in the remaining 85% of deductive studies,
and to what extent testing ad-hoc hypotheses not grounded in theory can advance our principled understanding of psychological phenomena.
These statistics suggest that theory has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely derived from theory,
and test results rarely contribute to the improvement of existing theories.
The paradoxical role of theory in psychology may help explain the observation that "theories either persist unchanged, immune to refutation,
or are forgotten" [check quote REF Meehl].

<!-- Scientific reform initiated by the open science movement has predominanty focused on improving deductive methods, overlooking the shortcomings of theory. -->
<!-- The present paper applies, for the first time, open science principles to theory. -->

## The Solution

Applying the FAIR principles to scientific theories facilitates transparent scholarly communication and accelerates cumulative knowledge acquisition for several reasons.

@@

One major obstacle to cumulative knowledge acquisition is the difficulty in finding relevant theories.
The primary unit of dissemination and search in psychology is often the academic paper.
A paper may contain multiple resources - including materials, data, code, and theory - but there is no unified search engine for theory,
nor standardized metadata or even a keyword to signal the presence of theory within a paper.
Making theories more findable would allow them to be systematically indexed, tagged, and made searchable.
This would allow researchers to easily identify relevant theories to inform their hypotheses,
grounding their work in established theoretical foundations.
Making theories findable also increases the impact and reuse potential of theories across scientific contexts.
For example, it becomes possible to conduct meta-research on theories, 
in the same way libraries and search engines have enabled scholars to study the literature via systematic reviews.
Moreover, theories might have transdisciplinary impact, either through direct application or through analogical modeling.
In analog modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
For example, predator-prey models have inspired theories of XXX, and the Eysenck model of atomic magnetism has inspired a network theory of depression.

One effort that improves theories' findability is post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website [REF GRAY].
Similarly, Borsboom and colleagues seek to establish a database of psychological theories [REF BORSBOOM].
Post-hoc curation does not address the root cause of the lack of Findability, however.
This would be achieved by modular publishing of theories as individually citable academic assets, with adequate metadata that is indexed in standardized repositories,
similar to the current practice of publishing empirical data in standardized repositories (e.g., DataVerse).
As with empirical data, these theories could still be connected to a specific paper which might serve as documentation and the canonical reference for the resource.

@@

A third, and related, impediment is a phenomenon we will call "dependency on the author" (DOA).
DOA occurs whenever a theory cannot be understood by independent scholars,
thus requiring the original author for interpretation and clarification.
We have heard DOA referred to apocryphally as the "ask Leon" phenomenon,
as graduate students were supposedly told to ask Leon Festinger to explain to them how their misconstrual of cognitive dissonance theory had caused their experiments to yield null results.
DOA is an obstacle when theory is insufficiently clear to derive hypotheses from it without first asking the author what they meant.
It also renders theories immune to refutation,
because the author can claim that the theory was misconstrued when confronted with falsifying evidence, thus making it a moving target.
DOA relates to the discourse on "Great Man Theorizing" [@guestWhatMakesGood2024] because dependency on the author creates a potential for gatekeeping - the author could insist that work requires their involvement or denounce work conducted outside their purview as illegitimate, which violates checks and balances of scientific research.

DOA is also a problem because authors may not understand their own theories well.
illustrated by instances where third parties identify logical inconsistencies within a theory [e.g., @kissnerIDENTIFICATIONLOGICALINCONSISTENCY2008].

@@

Interoperability pertains to the property of digital research artefacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016a].
As the original definition of interoperability was somewhat narrow and abstract,
the concept has recently been further refined in terms of facilitating "successful communication between machines and between humans and machines", where "A and B are considered X-interoperable if a common operation X exists that can be applied to both" [@vogtFAIR20Extending2024].
This definition raises an interesting question, namely *interoperable for what?*
Suitable answers for FAIR theory may be: this theory is X-interoperable for deriving testable null-hypotheses, or for the purpose of selecting relevant control variables,
or for the purpose of indicating the conditions necessary for observing a particular phenomenon.

These definitions suggest that FAIR theory should, ad minimum,
be instantiated in a human- and machine-readable file format, as previously recommended [@vanlissaWORCSWorkflowOpen2021].

@@

According to Kurt Lewin, "there's nothing as practical as a good theory" [@lewinPsychologyProcessGroup1943] - but as we argued bfore,
there are clear signs that theory rarely plays a practical role in the work of contemporary psychologists.
Interoperability makes it possible to integrate theory into scientists' day-to-day work in a practical way.
If a theory is X-interoperable with software used for data analysis,
it becomes possible to integrate theory directly into analysis workflows.
For example, interoperable theory could be used to derive machine-readable hypotheses [@lakensImprovingTransparencyFalsifiability2021],
to select control variables for causal inference [@cinelliCrashCourseGood2022],
and to preregister the inferential procedure that would lead to specific modifications of a theory after analyzing empirical data [@peikertReproducibleResearchTutorial2021]
Additionally, theories should be interoperable with each other:
for instance, it should be possible to embed a specific theory about the process of emotion regulation [e.g., @grossEmotionRegulationCurrent2015] within a theory of emotion regulation development [@morrisRoleFamilyContext2007].

@@

The ultimate goal of the FAIR principles is to facilitate the reuse digital research artefacts.
While theory is ostensibly held in high regard in the social sciences,
some have described a norm *against* theory reuse:
*"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's"* [@mischelToothbrushProblem2008].
Given that theory is the vehicle of cumulative knowledge acquisition [@degrootMethodologieGrondslagenVan1961],
such a norm impedes scientific progress.
Even if scholars wish to diverge substantially from prior theory,
explicitly refering back to it enables clear comparison of the differences [@ramGitCanFacilitate2013].
Reusing and adapting existing theories makes it possible to study the provenance of a theory and investigate how well different iterations of the theory explain emprical evidence [@vanlissaUsingEndpointsCheck2023]

In our workshops on FAIR Theory, we often hear questions such as "who owns theory",
and "who determines how a theory may be reused or changed"?
Licensing theories for reuse provides an unambiguous answer to such questions.
These questions suggest that there is a norm against modifying theory without consent from the author.
However, scholars might not be aware of the legal limits to the rights authors can claim over theory.
An important concept in copyright law is the *idea-expression dichotomy* [@bently2010copyright].

<!-- o. Ideas, facts, and concepts are not protected by  Although they are not protectable by copyright, the expression of those ideas, facts, and concepts are protectable, such as in a description, explanation, or illustration or as a database of facts. Separating an idea from the expression or manifestation of that idea is known in copyright law as the  -->
Copyright explicitly does not
*"extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery"*,
but it may extend to creative works expressing that idea (e.g., writing, visual illustrations).
Copyright pertaining to recipes offer an interesting perspective:
Whereas the recipes themselves are not protected, *"substantial literary expression in the form of an explanation or directions"* can be protected by copyright.
Applying these principles to FAIR theory, it seems that vague, ambiguous verbal explanations of theories are more likely to qualify for copyright protection than formal theories.
If copyright does not cover (formal) theory,
then it would be pointless to adopt a license that assumes copyright protection.
Furtermore, even if copyright would apply, academic research is covered under "fair use" exemptions to copyright law,
so copyright would impose few restrictions to theory reuse in scholarly communication.
Given these considerations, the CC0 (no rights reserved) license seems most appropriate for FAIR Theory;
it explicitly waives all rights and encourages reuse.
Simultaneously, to comply with the interoperability principle, the principle of comprehensive citation [@aalbersbergMakingScienceTransparent2018],
and the spirit of good scholarship,
it is essential that scholars do comprehensively cite theory,
including prior work that new theories are based on,
even in absence of legal obligations to do so.

3. Interoperability: Facilitating Integration Across Fields
In many scientific disciplines, researchers often operate in isolated silos, using different terminologies, methodologies, and conceptual frameworks. The interoperability aspect of FAIR focuses on the ability of data (or in this case, theories) to be integrated with other systems using common standards, formats, and vocabularies.

How it helps: If theories are made interoperable, they can be used in multi-disciplinary research and can better interface with diverse fields. This reduces duplication of effort and enables researchers from different areas to combine and compare theories, bridging gaps between disciplines and fostering innovative, cross-disciplinary approaches. Researchers can easily synthesize theoretical perspectives, generating new hypotheses that are more comprehensive and grounded in a wider range of evidence.
4. Reusability: Enabling Continuous Improvement of Theories
For knowledge to accumulate meaningfully, theories need to be adaptable and subject to continuous revision as new data emerge. The reusability principle calls for thorough documentation and community standards to ensure that data (or theories) can be applied in new contexts.

How it helps: Reusable theories would provide clear metadata, version histories, and supporting evidence, allowing other researchers to test, refine, or extend them as new empirical findings emerge. This improves the feedback loop between theory and experimentation, ensuring that test results inform future theoretical development. It encourages the use of theories not just for their original purpose but in new and evolving contexts, thus ensuring that the theory remains dynamic and continuously refined over time.
Conclusion: Making Theories FAIR Accelerates Scientific Progress
Adopting the FAIR principles for theories can address key challenges in the current research landscape, where theories often remain isolated and underutilized. By making theories findable, accessible, interoperable, and reusable, researchers can ensure that their work is grounded in a shared, transparent, and cumulative body of knowledge. This approach enhances scholarly communication, allowing for greater scrutiny, replication, and collaboration across disciplines, ultimately leading to faster, more reliable, and more impactful scientific progress.

@@

The FAIR Principles were devised to make scholarly data more findable, accessible, interoperable, and reusable.
From their inception, these principles were developed with "other research resources" in mind.
Scholars have translated the FAIR principles to, e.g., research software [REF Lamprecht].
The present paper further extends the FAIR principles' definition to theory, see Table \@ref(tab:tabfair).
We establish guiding principles for instantiating theory as a FAIR digital research artifact,
and provide applied examples to encourage their adoption.

@@

In the spirit of DORA, extending the FAIR principles to theory helps researchers obtain credit for their theoretical contributions - obviating the necessity of publishing a theoretical paper, which can be challenging.
From a meta-science perspective, FAIR theory facilitates studying the state of theory in a particular subfield, and comparing theories' substantive and structural properties. Version control and cross-referencing additionally enable tracing and studying the ancestry and development of theories.

FAIR theory provides a clear deliverable, and a clear goal, for scholars and institutions seeking to promote contributions to theory.

There are key distinctions between theory and other FAIR digital research artefacts. With this in mind, following the example of Lamprecht and colleagues, we reflect on how the criteria underlying the FAIR Principles apply to theory.

@@

## Version Control

* One source of potential improvements of theory methodology that has not been previously considered is computer science.
* The process of "iteratively improving" digital objects - in this case, computer code - is well understood.
* Recent work like the FAIR software principles has demonstrated that ideals of open science apply to computer science as well.
* This paper argues that, conversely, principles of computer science - particularly version control, algorithmic hypothesis generation (find better word; this is about using the digital theory object to derive implied hypotheses), and integrated testing, can also be used to improve theory methods in the social science.
* We introduce "FAIR theory", a digital research artifact to represent formal social scientific theories
* FAIR theory can be version controlled; any time new insights require modifications of the theory, these modifications can be documented in a traceable and reversable manner. Version control also enables diffuse collaboration in theory development, as other researchers can submit "pull requests" to suggest modifications of a theory, or can "fork" existing theories to create a spin-off from an existing theory.

>>>>

abstract: |
  Test test.
keywords          : "meta theory, theory formation, cumulative science, formal models"
wordcount         : "`r tryCatch(wordcountaddin::word_count(here::here('manuscript.Rmd')))`"
bibliography      : "theory-specification.bib"
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
---

@@

The FAIR Guiding Principles (hereafter: FAIR principles) were established to make research data more Findable, Accessible, Interoperable and Reusable [REF].
Since their inception, scholars have demonstrated their relevance for making other information artefacts more open.
This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
To this end, we introduce "FAIR theory":
a digital representation of theory, compliant with the FAIR principles.
By improving the efficiency of scholarly communication, FAIR Theory has the potential to accelerate cumulative knowledge acquisition and advance social scientific research.

@@

Scholars have raised concerns about the state of theory in social science for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
Two main concerns are that, first, social scientific theories lack precision and clarity compared to theories in the physical sciences [@szollosiArrestedTheoryDevelopment2021].
In other words, social scientific theories lack *formalization* .
A second concern is the lack of transparent and democratic scholarly communication about psychological theory.

Given these concerns, it is unfortunate that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processes of theory construction and -improvement have been largely overlooked.
The present paper restores balance by applying, for the first time,
open science principles to psychological theory.
We apply the FAIR principles to scientific theories,
introducing the concept of *FAIR Theory* to 
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

In a progressive research program [@lakatosHistoryScienceIts1971],
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because deductive research is over-represented in the literature: 
According to one estimate, 89.6% of paper published in psychology tests hypotheses [@kuhbergerPublicationBiasPsychology2014].
Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
Only 15% of deductive studies referenced any theory, and theory was often not cited in connection to the hypothesis [@mcphetresDecadeTheoryReflected2021].
The remaining 85% of deductive studies lacked an explicit derivation chain from theory to hypothesis.
Perhaps such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is important to make these explicit [REF Rasoul].
Or, perhaps the hypotheses are of no substantive interest [REF Hoijtink], and researchers are simply testing them as part of a cultural ritual [REF Gigerenzer].
No matter what the explanation - testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.

According to the aforementioned statistics, theory has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely derived from theory,
and test results do not routinely contribute to the improvement of existing theories.
The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology "lack the cumulative character of scientific knowledge. They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest" [@meehlTheoreticalRisksTabular1978].

To date, scientific reform initiated by the open science movement has predominanty focused on improving deductive methods, overlooking the shortcomings of theory.
The present paper applies, for the first time, open science principles to theory.
Applying the FAIR principles to scientific theories facilitates transparent scholarly communication and accelerates cumulative knowledge acquisition.

@@

One major obstacle to cumulative knowledge acquisition is the difficulty in finding relevant theories.
One contributing cause of theories' lack of Findability is the lack of standardized metadata or even a keyword to signal the presence of theory within a paper, and - consequently - the absence of a unified search engine for theory.
Making theories more findable through rich metadata would allow them to be systematically indexed, tagged, and made searchable.
This would allow researchers to easily identify relevant theories to inform their hypotheses,
grounding their work in established theoretical foundations.
Making theories findable also increases the impact and reuse potential of theories across scientific contexts.
For example, it becomes possible to conduct meta-research on theories, 
in the same way libraries and search engines have enabled scholars to study the literature via systematic reviews.
Moreover, theories might have transdisciplinary impact, either through direct application or through analogical modeling.
In analog modeling, the structure of a theory from one discipline is applied to a phenomenon in another field.
For example, predator-prey models have inspired theories of XXX, and the Eysenck model of atomic magnetism has inspired a network theory of depression.

Another contributing cause to the lack of Findability is that the primary unit of dissemination and search in psychology is still the academic paper.
A paper may contain multiple resources - such as materials, data, code, and theory - but if these are not merely described in text, and not instantiated as separate informational artefacts, their findability is limited.
This would be achieved by modular publishing of theories as individually citable academic assets, with adequate metadata that is indexed in standardized repositories,
similar to the current practice of publishing empirical data in standardized repositories (e.g., DataVerse).
As with empirical data, these theories could still be connected to a specific paper which might serve as documentation and the canonical reference for the resource.

There have been notable efforts to improve theories' findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website [REF GRAY].
Similarly, Borsboom and colleagues seek to establish a database of psychological theories [REF BORSBOOM].
Post-hoc curation does not address the root cause of the lack of Findability, however.
Ideally, Findability would be addressed ante-hoc, through modular publishing and documentation with rich metadata.

@@

A third impediment arises when theories have a "dependency on the author" (DOA).
DOA occurs when a theory cannot be understood by independent scholars,
thus requiring the original author for interpretation and clarification.
We have heard DOA referred to apocryphally as the "ask Leon" phenomenon,
as graduate students were supposedly told to ask Leon Festinger to explain to them how their misconstrual of cognitive dissonance theory had caused their experiments to yield null results.
DOA relates to the discourse on "Great Man Theorizing" [@guestWhatMakesGood2024] because it enables gatekeeping: an author could insist that work requires their involvement or denounce work conducted outside their purview as illegitimate,
which violates checks and balances of scientific research.
DOA also renders theories immune to refutation,
because the author can claim that the theory was misconstrued when confronted with falsifying evidence, thus making it a moving target  [@szollosiArrestedTheoryDevelopment2021].
DOA is inherently problematic because authors may not understand their own theories well.
This is illustrated by cases where third parties identify logical inconsistencies within a theory [e.g., @kissnerIDENTIFICATIONLOGICALINCONSISTENCY2008].
In short, DOA unduly impedes scientific progress, and authors should make good-faith efforts to make theories as accessible as possible; both in terms of availability and in terms of interpretability.

@@

Interoperability pertains to the property of information artefacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016a].
As the original definition of interoperability was somewhat narrow and abstract,
the concept has recently been further refined in terms of facilitating "successful communication between machines and between humans and machines", where "A and B are considered X-interoperable if a common operation X exists that can be applied to both" [@vogtFAIR20Extending2024].
This definition invites the question: *interoperable for what purpose?*
Suitable answers for FAIR theory may be:
this theory is X-interoperable for deriving testable null-hypotheses,
or for the purpose of selecting relevant control variables,
or for the purpose of indicating the conditions necessary for observing a particular phenomenon.
These definitions suggest that FAIR theory should, ad minimum,
be instantiated in a human- and machine-readable file format, as previously recommended [@vanlissaWORCSWorkflowOpen2021].
Furthermore, theories may have specific properties that incur affordances in terms of X-interoperability;
for example, Table \@ref(tab:tabmeehl) illustrates the affordances of Meehl's nine properties of strong theories.

```{r}
data.frame(
  "Property" = c("1) Ontology", "2) Causal connections", "3-8) Functional Form", "9) Numerical Value"),
  "X-interoperability" = c("Variable selection", "Model specification, covariate selection, causal inference", "Deriving specific hypotheses", "Simulating data"), check.names = FALSE
) |>
papaja::apa_table()
```

@@

Kurt Lewin's adage "there's nothing as practical as a good theory" [@lewinPsychologyProcessGroup1943] seems to imply that theory plays a practical role in the day-to-day work of psychological researchers.
But, as we argued before, there are clear signs that this is not the case.
The examples of X-interoperability offered in Table \@ref(tab:tabmeehl) illustrate that much can be gained by integrating theory directly into analysis workflows, and by making theory X-interoperable within software used for analysis.
For example, interoperable theory could be used
to select control variables for causal inference [@cinelliCrashCourseGood2022],
or to preregister the inferential procedure that would lead to specific modifications of a theory after analyzing empirical data [@peikertReproducibleResearchTutorial2021],
or to derive machine-readable hypotheses [@lakensImprovingTransparencyFalsifiability2021] which could be automatically evaluated through integration testing [@vanlissaUsingEndpointsCheck2023].
Furthermore, theories can be X-interoperable with each other to enable nesting, or using one theory to clarify elements of another theory.
For example, it should be possible to embed a theory about emotion regulation [e.g., @grossEmotionRegulationCurrent2015] within a theory of emotion regulation development [@morrisRoleFamilyContext2007].

@@

Some have argued that there is a norm *against* theory reuse:
*"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's"* [@mischelToothbrushProblem2008].
Such a norm impedes scientific progress.
Cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data [@degrootMethodologieGrondslagenVan1961].
Theories are made reusable by thorough documentation,
appropriate licensing, 
and detailed provenance (e.g., through version control, cross-referencing, and semantic versioning).

In our workshops on FAIR Theory, we often hear questions such as "who owns theory",
and "who determines how a theory may be reused or changed"?
Licensing theories for reuse provides an unambiguous answer to such questions.
These questions imply a norm against modifying theory without consent from the author reminiscent of the aforementioned problem of dependency on the author.
However, copyright law limits the rights authors can claim over theory based on the *idea-expression dichotomy* [@bently2010copyright],
<!-- o. Ideas, facts, and concepts are not protected by  Although they are not protectable by copyright, the expression of those ideas, facts, and concepts are protectable, such as in a description, explanation, or illustration or as a database of facts. Separating an idea from the expression or manifestation of that idea is known in copyright law as the  -->
which holds that copyright explicitly does not
*"extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery"*.
It may, however, extend to creative works expressing that idea (e.g., writing, visual illustrations).
<!-- Copyright pertaining to recipes offer an interesting perspective: -->
<!-- Whereas the recipes themselves are not protected, *"substantial literary expression in the form of an explanation or directions"* can be protected by copyright. -->
Applying these principles to FAIR theory, it seems that vague, ambiguous verbal explanations of theories - in other words, those that fall short in terms of the Accessibility criterion - are more likely to qualify for copyright protection than formal theories.
If copyright limits Reusability and does not cover ideas in their purest form (like formal theory),
then it might be counterproductive and possibly misleading to adopt a license that assumes copyright protection.
Furthermore, even if copyright would apply, academic research is covered under "fair use" exemptions,
so copyright would pose few restrictions to Reusability in scholarly communication.
Given these considerations, the CC0 (no rights reserved) license seems most appropriate for FAIR Theory;
it explicitly waives all rights and encourages reuse.
In principle, CC0 does not require attribution.
Nevertheless, is essential that scholars do comprehensively cite theory,
including prior work that new theories are based on,
even in absence of legal obligations to do so,
to meet the definition of Reusability (R1.2, Table \@ref(tab:tabfair) and to comply with other definitions of open scholarship [@aalbersbergMakingScienceTransparent2018].

We can take inspiration from the field of computer science for well-established processes for iteratively improving information artefacts,  like computer code.
Using version control systems, like Git, would enhance the reusability of FAIR theory by thoroughly documenting every modification in a traceable and reversible manner.
Git also facilitates diffuse and adversarial collaboration,
as independent researchers can create independent versions of existing theories through "forking", or suggest modifications to existing theories via "pull requests".
In sum, version control using Git enables systematic, collaborative, and transparent theory development,
enables studying the provenance of a theory and investigating how well different iterations of the theory explain empirical evidence [@vanlissaUsingEndpointsCheck2023].

Even if scholars wish to diverge substantially from prior theory,
explicitly referring back to it enables clear comparison of the differences [@ramGitCanFacilitate2013].

From a meta-science perspective, FAIR theory facilitates studying the state of theory in a particular subfield, and comparing theories' substantive and structural properties.

Making Theories FAIR Accelerates Scientific Progress
Adopting the FAIR principles for theories can address key challenges in the current research landscape, where theories often remain isolated and underutilized. By making theories findable, accessible, interoperable, and reusable, researchers can ensure that their work is grounded in a shared, transparent, and cumulative body of knowledge. This approach enhances scholarly communication, allowing for greater scrutiny, replication, and collaboration across disciplines, ultimately leading to faster, more reliable, and more impactful scientific progress.

@@

The FAIR Principles were devised to make scholarly data more findable, accessible, interoperable, and reusable.
From their inception, these principles were developed with "other research resources" in mind.
Scholars have adapted the FAIR principles to, e.g., research software [REF Lamprecht].
Yet there are key distinctions between theory and other FAIR information artefacts.
With this in mind, the present paper further extends the FAIR principles' definition to theory, see Table \@ref(tab:tabfair).
We establish guiding principles for instantiating theory as a FAIR digital research artifact,
and provide applied examples to encourage their adoption.

@@

FAIR theory provides a clear deliverable, and a clear goal, for scholars and institutions seeking to promote contributions to theory.
In the spirit of DORA, this helps researchers obtain credit for their theoretical contributions - obviating the necessity of publishing a purely theoretical paper, which can be challenging.

@@
